---
title: "Project 1"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(colorspace)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)

olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')

# data prep:
olympics_top <- olympics %>%
  mutate(team = ifelse(team == "Soviet Union", "USSR/Russia", team)) %>%
  mutate(medal = ifelse(is.na(medal), "no medal", medal)) %>%
  mutate(gold = ifelse(medal == "Gold", "gold medalist", "no gold")) %>%
  mutate(medalist = ifelse(medal != "no medal", "medalist", "no medal")) %>%
  mutate(decade = year - (year %% 10)) %>%
  group_by(team) %>%
  mutate(count = n()) %>%
  filter(count > 5500) %>%
  select(-count) %>%
  ungroup() %>%
  group_by(sport, season) %>%
  mutate(count = n()) %>%
  filter(count > 700) %>%
  select(-count) %>%
  ungroup() %>%
  filter(!sport %in% c("Art Competitions", "Handball", "Short Track Speed Skating", "Modern Pentathlon", "Shooting", "Canoeing", "Athletics", "Hockey", "Water Polo"))
```

We will work with the dataset `olympics_top` that contains data for the Olympic Games from Athens 1896 to Rio 2016 and has been derived from the `olympics` dataset. More information about the dataset can be found at: https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-07-27/readme.md
The dataset, `olympics_top`, contains four new columns: `decade` (the decade during which the Olympics took place), `gold` (whether or not the athlete won a gold medal), `medalist` (whether or not the athlete won any medal) and `medal` (if the athlete won "Gold", "Silver", "Bronze" or received "no medal").

### **Part 1**

**Question:** Which sports have the tallest or shortest athletes? And does the distribution of heights change for the various sports between medalists and non-medalists? 

We recommend you use box plots for the first part of the question and use a ridgeline plot for the second part of the question. 

**Hints:**

- To order boxplots by the median, you may have add the following to your ordering function to remove missing values before ordering: `na.rm = TRUE`

- To trim the tails in your ridgeline plot, you can set 
`rel_min_height = 0.01` inside `geom_density_ridges()`.

**Introduction:** 

*The dataset* `olympics_top` *is a historical dataset containing information on Olympic games competitors from 1896 through Summer 2016. Variables include information on age, weight, height, sex, where the games were held, what event the athlete competed in, if they received a medal, and so on. This information is encoded long, with each observation encoding information about a particular athlete's performance in a particular Olympic games. Data was scraped from www.sports-references.com by a Kaggle user, after original compilation by sports hobbyists.* 

*To answer the question above, I'll need the information on sport, sex, height, and medalist status. I'll also include age and weight in some regression models, since those would likely be confounding variables and introduce bias if omitted.*

```{r}
head(olympics_top)
# Lets see what's in the data...
table(olympics_top$games)
# games is encoded as a character string, so we're going to need to generate a table to get an idea about the range of values.
```

**Approach:** 

*I'll answer the first part of the question by calling* `summary()` *on height, grouped by sport and sex. I'll then use the* `geom_boxplot()` *feature of* `ggplot()` *to visualize my findings. For the second part of the question, I'll run a* `lm()` *of height, age and weight on medalist status, grouped by sport and sex. I'll also run regressions excluding age and weight, and test for correlation on the same parameters. Finally, I'll use a* `geom_density_ridges()` *plot to visualize my findings.*

**Analysis:**

```{r}
# part one, summary calls
olympics_top %>%
  group_by(sport, sex) %>%
  summary(height)

#part one, boxplot()
ggplot(olympics_top, aes(reorder(sport, height, na.rm = TRUE), height, fill = sex)) +
  geom_boxplot(position = position_dodge(width = .5), width = .5) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

```{r}
# part two, lm()

for(i in olympics_top$name) {
    olympics_top$name <- 
  }
lm_noctrls <- lm(height ~ )

# part two, ridgeline()


```


**Discussion:** 

*Your discussion of results here.*

### **Part 2**

**Question:** *Your question here.*

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**
```{r }
# Your R code here
```

```{r }
# Your R code here
```

**Discussion:** *Your discussion of results here.*
